Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-01-19 paper/sheaf_lics.tex
minor changes

As a prerequisite, we encourage the reader be familiar with type theory and
in particular the point of view developed
in~\cite{hottbook}. Nevertheless, we recall most of the central definitions 
that we use so that the paper is sufficiently self-contained.
%

@@

Also proposition truncation could be seen as a primitive type former
in homotopy type theory, it can be defined using higher inductive
types. We adopt this point of view as it allows to maintain the slogan
that homotopy type theory is type theory plus univalence and higher
inductive types.
%

@@

In our formalization of sheafification defined
Section~\ref{sec:sheafification} (done in 
\texttt{sheaf\_induction.v}), two very technical facts required to
show that $\separated$ is a modality have been
admitted. 
%
It is due to a lack of important lemmas on paths between paths in
the Coq/HoTT library.
%
Nevertheless, we are confident that with enrichment of the Coq/HoTT
library, they can be proven and have been admitted only by lack of
time.
%
In the proof that diagrams are equal in lemma~\ref{lem:sepiscolim},
the equality of projections is admitted, because if this proof is
obvious on paper, it becomes really technical when formalized.
%
This suggest that we should try to fine a better way to encode
projections, with a better computational content.
%
As a consequence, the fact that $P$ is a colimit of the diagonal
closed diagram is admitted too, but again, this proof should be really
short as it is on paper.

@@

In this paper, we have defined a way to extend homotopy type theory
with new principles, leveraging Lawvere-Tierney sheafification to
higher topos theory. 
%
Using sheafification, it is possible to add new logical principles,
like the law of excluded middle, to homotopy type theory and this
without axiom.
%
Beyond the result, our work is part of a more general line of works
which aim to illustrate that homotopy type theory is a very promising
theory for formalizing mathematics inside a proof assistant.

In future works, we would like to improve this construction in two
ways:
\begin{itemize}
\item 
  At the moment, sheafification only works with truncated
  types. Also this is fine from a programming point of view, 
  many important types in homotopy theory are not
  truncated, like the circle $\mathbb S^1$.  Thus, we plan to extend
  the sheafification to all types, even if non truncated by using a
  limit construction.
\item 
  In this work, a particular case of a Giraud-Rezk-Lurie axiom has
  been admitted. The work done in~\cite[Chapter 6]{lurie} suggests,
  considering that homotopy type theory is a language for higher topos
  theory, that this axiom is provable. We plan to investigate if the
  higher topos theoretic proof can be adapted to our setting. 
  %
  % A computer-checked proof of this fact would improve our
  % developpement.
\end{itemize}
Moreover, in topos theory, Lawvere-Tierney subsumes Grothendieck
(\cite[Section~V.4]{maclanemoerdijk}) in the sense that any
Grothendieck topology gives rise to a Lawvere-Tierney topology with
the same notion of sheaves. Higher Lawvere-Tierney sheaves are
presented here, and higher Grothendieck sheaves have been defined
in~\cite{lurie}. It should be interesting to check if the subsumption
remains true in higher topos theory.

>>>>

As a prerequisite, we encourage the reader to be familiar with type theory and
in particular the point of view developed
in~\cite{hottbook}. Nevertheless, we recall most of the central definitions 
that we use so that the paper is sufficiently self-contained.
%
We note $\prod_{x:T} U$ for the dependent product, $\sum_{x:T} U$ for
the dependent sum, and $\pi_1, \pi_2$ for the first and second
projection of a dependent pair. We use informal mathematical language
instead of type theory whenever it is possible to ease the
reading, without making our statement imprecise. In particular,
(higher) inductive types are defined using itemization to avoid an
overhead of notation.

@@

Also proposition truncation could be seen as a primitive type former
in homotopy type theory, it can be defined using higher inductive
types~\cite{lumsdaine2011higher,lumsdaine2013higher}. We adopt this
point of view as it allows to maintain the slogan that homotopy type
theory is type theory plus univalence and higher inductive types.
%

@@

In our formalization of sheafification defined in
Section~\ref{sec:sheafification} (done in 
\texttt{sheaf\_induction.v}), two very technical facts required to
show that $\separated$ is a modality have been
admitted. 
%
It is due to a lack of technical lemmas on paths between paths in
the Coq/HoTT library.
%
Nevertheless, we are confident that with an enrichment of the Coq/HoTT
library, they can be proven and have been admitted only by lack of
time.
%
In the proof that diagrams are equal in lemma~\ref{lem:sepiscolim},
the equality of projections is admitted, because if this proof is
obvious on paper, it becomes really technical when formalized.
%
This suggest that we should try to fine a better way to encode
projections, with a better computational content.
%
As a consequence, the fact that $P$ is a colimit of the diagonal
closed diagram is admitted too, but again, this proof should be really
short as it is on paper.

@@

In this paper, we have defined a way to extend homotopy type theory
with new principles, leveraging Lawvere-Tierney sheafification to
higher topos theory. 
%
Using sheafification, it is possible to add new logical principles to
homotopy type theory, like the law of excluded middle, and this
without axiom.
%
Beyond this result, our work is part of a more general line of works
which aim to illustrate that homotopy type theory is a very promising
theory for formalizing mathematics inside a proof assistant.

As a future work, we would like to improve this construction in two
ways:
\begin{itemize}
\item 
  At the moment, sheafification only works with truncated
  types. This is fine from a programming point of view, 
  but many important types in homotopy theory are not
  truncated, \eg the circle $\mathbb S^1$.  Thus, we plan to extend
  the sheafification to all types by using a
  limit construction.
\item 
  In this work, a particular case of a Giraud-Rezk-Lurie axiom has
  been admitted. The work done in~\cite[Chapter 6]{lurie} suggests,
  considering that homotopy type theory is a language for higher topos
  theory, that this axiom is provable. We plan to investigate whether the
  higher topos theoretic proof can be adapted to our setting.
  %
  % A computer-checked proof of this fact would improve our
  % developpement.
\end{itemize}
Moreover, in topos theory, Lawvere-Tierney subsumes Grothendieck~\cite[Section~V.4]{maclanemoerdijk} in the sense that any
Grothendieck topology gives rise to a Lawvere-Tierney topology with
the same notion of sheaves. Higher Lawvere-Tierney sheaves are
presented here, and higher Grothendieck sheaves have been defined
in~\cite{lurie}. It should be interesting to check if the subsumption
remains true in higher topos theory.
